A Match Made in Landfills?

### 2023-05-05 5 Manuscript/manuscript/Manuscript.qmd
Manuscript edits

In 2019, the USGS revealed that more than half of GPS-tracked gulls (*Larus* ssp.) had acquired antimicrobial resistant *E. coli* from landfills and disseminated it long distances to pristine, unaltered habitats [@Ahlstrom2019]. This supports the belief that birds who utilize and move between human-dominated and natural landscapes may become both reservoirs and efficient transporters of resistance determinants to natural environments.For wildlife in general, resistance prevalence tends to increase with proximity to human populations [@Niu2013]. It is believed that animals that rely on anthropogenic resources are often concentrated at high densities that promote the large-scale mixing of bacteria, encouranging horizontal gene transfer (HGT) facilitating the rise of novel resistant strains [@Wellington2013]. For instance, birds that encounter human waste (e.g., by drinking contaminated water) can acquire resistance genes through the exchange of naturally present genes in the bacteria they carry, with the genes in the waste---this can co-select for mobile genetic elements carrying multiple resistant genes [@Wellington2013]. Moreover, ecological factors such as migration and high densities during breeding season may increase acquisition rates and the dissemination of AMR . The western white stork are one such species that heavily utilizes agricultural and urbam areas, providing a unique opportunity to explore how anthropogenic land use influences the carriage of antimicrobial resistance genes.

@@

Evaluate both antimicrobial gene resistence diversity and burden in white storks that utilize landfills and natural areas over the breeding season of years 2020 to 2021.

@@

We have 126 observations taken over a period of 2 years. We are evaluating for the presence and burden of antimicrobial resistance genes in white stork feces. Codebook is a WIP located in raw data folder.

@@

Binomial generalized linear mixed models (GLMMs) were used to predict multi-drug resistance (MDR) in white storks, with nest identification classified as a random effects parameter. Of these, the highest performing model contained the predictors landfill use index (LUI) and age (Table 1). In this model, an increase in the age of the bird was associated with a higher likelihood of MDR being present (Table 2), while an increase in LUI was found to be associated with a decreased likelihood of MDR presence. A second competitive model, with a 97.20% performance score, contained only age as the predictor.In the analysis, several models where overfit, not including the top two models. However, this is likely due to the small sample size and the restricted range of the standardized variables.

@@

Linear mixed models (LMMs) were used to predict total antimicrobial gene burden in white storks, with nest identification again classified as a random effects parameter. The global model, containing the variables landfill use index, age, sample period, and nest success, was the highest performing model (Table 3). In this model, each predictor variable was positively correlated with antimicrobial gene burden (Table 4). The next highest performing model contained age as a single predictor, with a model performance score of 86.39%. Similarly to the top analysis, several models where overfit likely due to similar reasons.

@@

-   As multi-drug resistance increases throughout the breeding season, mean LUI decreases and nest success decreases. There is a positive effect between increased LUI and nest success [@Lopez-Garcia2021]. Intuitively this makes some sense due to MDR being associated with increased anthropogenic exposure, heavy metal exposure, and increased virulence of pathogenic bacteria. More target research would be needed to better understand how MDR may effect nest success long-term.

-   Resistance gene burden appears to increase as mean LUI decreases and age increases. Most of the burden is due to blaTEM, a common resistance gene associated with beta-lactam resistance. Beta-lactam resistance is one of the most studied AMRs in wildlife. Drugs within the Beta-lactam class are often first line antibiotics, such as penicillin, and are active against many gram-positive, gram-negative, and anaerobic organisms. At this point, beta-lactam resistance is thought to be prevalent on most landscapes due to human or livestock encroachment thus is often used as a measure of anthropogenic impact. While sampling periods did not explain burden, the top blaTEM model did show a trend in burden increasing from the 2nd sampling period to the 3rd sampling period.This may be related to shear increase of foraging required to feed ~3 to 4 kilogram fledglings.This foraging could have occurred in areas other then the landfill, such as neighboring cattle pastures which may also be sources for ARGs.

@@

-   While none of our models resulted in statistically significant data, valuable trends can be seen in respect to the relationship between breeding season events, AMR trends, and landfill use. Hopefully, these trends can foster future research to consider expandings beyond the hypothesis that landfills are the risky place for wildlife to acquire ARGs.

-   The dataset utilized for the models was much smaller then previously envisioned due to COVID-19 barring landfill data collection throughout the entirety of 2020. Perhaps with more observations, the results of the models may have been stronger and variables may have been more normally distributed. Moving forward, it may be best to transform our standardized variables using a log transformation to attempt to achieve a more normal distribution for analysis, especially in regard to our GLMMs.

>>>>

In 2019, the USGS revealed that more than half of GPS-tracked gulls (*Larus* ssp.) had acquired antimicrobial resistant *E. coli* from landfills and disseminated it long distances to pristine, unaltered habitats [@Ahlstrom2019]. This supports the belief that birds who utilize and move between human-dominated and natural landscapes may become both reservoirs and efficient transporters of resistance determinants to natural environments.For wildlife in general, resistance prevalence tends to increase with proximity to human populations [@Niu2013]. It is believed that animals that rely on anthropogenic resources are often concentrated at high densities that promote the large-scale mixing of bacteria, encouraging horizontal gene transfer (HGT) facilitating the rise of novel resistant strains [@Wellington2013]. For instance, birds that encounter human waste (e.g., by drinking contaminated water) can acquire resistance genes through the exchange of naturally present genes in the bacteria they carry, with the genes in the waste---this can co-select for mobile genetic elements carrying multiple resistant genes [@Wellington2013]. Moreover, ecological factors such as migration and high densities during breeding season may increase acquisition rates and the dissemination of AMR . The western white stork are one such species that heavily utilizes agricultural and urbam areas, providing a unique opportunity to explore how anthropogenic land use influences the carriage of antimicrobial resistance genes.

@@

Evaluate both antimicrobial gene resistance diversity and burden in white storks that utilize landfills and natural areas over the breeding season of years 2020 to 2021.

@@

We have 126 observations taken over a period of 2 years. We are evaluating for the presence and burden of antimicrobial resistance genes in white stork feces. A codebook containing descriptions of variables and their abbreviations can be found in the raw data folder, located within the project repository.

@@

Binomial generalized linear mixed models (GLMMs) were used to predict multi-drug resistance (MDR) in white storks, with nest identification classified as a random effects parameter. Of these, the highest performing model contained the predictors landfill use index (LUI) and age (Table 1). In this model, an increase in the age of the bird was associated with a higher likelihood of MDR being present (Table 2), while an increase in LUI was found to be associated with a decreased likelihood of MDR presence. A second competitive model, with a 97.20% performance score, contained only age as the predictor. In this analysis, several models where overfit, not including the top two models. However, this is likely due to the small sample size and the restricted range of the standardized variables.

@@

Linear mixed models (LMMs) were used to predict total antimicrobial gene burden in white storks, with nest identification again classified as a random effects parameter. The global model, containing the variables landfill use index, age, sample period, and nest success, was the highest performing model (Table 3). In this model, each predictor variable was positively correlated with antimicrobial gene burden (Table 4). The next highest performing model contained age as a single predictor, with a model performance score of 86.39%. As with the GLMM analysis, several models, including the top model, where overfit, likely due to similar reasons.

@@

-   As multi-drug resistance increases throughout the breeding season, mean LUI decreases and nest success decreases. There is a positive effect between increased LUI and nest success [@Lopez-Garcia2021]. Intuitively, this makes some sense due to MDR being associated with increased anthropogenic exposure, heavy metal exposure, and increased virulence of pathogenic bacteria. More targeted research would be needed to better understand how MDR may effect nest success long-term.

-   Resistance gene burden appears to increase as mean LUI decreases and age increases. Most of the total burden is attributed to blaTEM, a common resistance gene associated with beta-lactam resistance. Beta-lactam resistance is one of the most studied AMRs in wildlife. Drugs within the Beta-lactam class are often first-line antibiotics, such as penicillin, and are active against many gram-positive, gram-negative, and anaerobic organisms. At this point, beta-lactam resistance is thought to be prevalent on most landscapes due to human or livestock encroachment and thus is often used as a measure of anthropogenic impact. While sampling periods did not explain burden, the top blaTEM model did show a trend in burden increasing from the 2nd sampling period to the 3rd sampling period.This may be related to sheer increase of foraging required to feed \~3 to 4 kilogram fledglings.This foraging could have occurred in areas other then the landfill, such as neighboring cattle pastures which may also be sources for ARGs.

@@

-   While none of our models resulted in statistically significant data, valuable trends can be seen in respect to the relationship between breeding season events, AMR trends, and landfill use. Hopefully, these trends can foster future research to consider expanding beyond the hypothesis that landfills are the risky place for wildlife to acquire ARGs.

-   The dataset utilized for the models was much smaller then previously envisioned due to COVID-19 barring landfill data collection throughout the entirety of 2020. Perhaps with more observations, the results of the models may have been stronger and variables may have been more normally distributed. While model residuals should be normally-distributed, model performance metrics indicated polarization in the residuals in several of our top model, which can lead to under- and over-fitting data. Moving forward, it may be best to transform our standardized variables using a log transformation to attempt to achieve a more normal distribution for analysis, especially in regard to our GLMMs.
